%version of 04-11-18

\chapter{TECHNIQUES FOR ``DOING'' MATHEMATICS}
\label{ch:doingmath}

\section{Reasoning via Rigorous Proof}
\label{sec:reasoning-via-proofs}

We survey some important proof techniques

\subsection{Proof, Implication, Equivalence}
\label{sec:basic-logic}



\medskip



\subsection{Proof by (Finite) Induction}
\label{sec:Induction}

\subsubsection{The proof technique}


\subsubsection{Sample proofs: verifying summation formulas}
\label{sec:Proof-Induction}


We illustrate the proof technique of (Finite) Induction by proving the
correctness of two familiar summation formulas: (1) the sum of the
first $n$ positive integers and (2) the sum of the first $n$ odd
positive integers.

\begin{prop}
\label{thm:sum-1-to-n-induction}
For all $n \in \N$,
\begin{eqnarray}
\nonumber
S_n \ \eqdef \ \sum_{i=1}^n \ i
 & \eqdef &
 1 + 2 + \cdots + (n-1) + n \\
\label{eq:sum-first-n}
 & = & {1 \over 2} n (n+1) \\
\nonumber
 & = & {{n+1}  \choose 2}.
\end{eqnarray}
\end{prop}

\begin{proof}
For every positive integer $m$, let {\bf P}$(m)$ be the proposition
\[  1 + 2 + \cdot + m \ = \ {{m+1} \choose 2}. \]
Let us proceed according to the standard format of an inductive
argument.

{\bf 1.} Because ${\displaystyle {2 \choose 2}} = 1$, proposition {\bf
  P}$(1)$ is true.

{\bf 2.} Let us assume, for the sake of induction, that proposition
{\bf P}$(m)$ is true for all positive integers strictly smaller than
$n$.  In particular, then, {\bf P}$(n-1)$ is true.

{\bf 3.} Consider now the summation
\[ 1 + 2 + \cdots + (n-1) + n. \]
Because {\bf P}$(n-1)$ is true, we know that
\[ 1 + 2 + \cdots + (n-1) \ = \ {n \choose 2}.  \]
By direct calculation, we see that
\begin{eqnarray*}
{n \choose 2} + n
  & = & \frac{n(n-1)}{2}  \ + \ n \\ 
  & = & \frac{n^2 - n + 2n}{2} \\
  & = & \frac{n^2 + n}{2} \\
  & = & {{n+1} \choose 2}
\end{eqnarray*}

Because $n$ is an arbitrary positive integer, we conclude that
{\bf P}$(n)$ is true whenever
\begin{itemize}
\item
{\bf P}$(1)$ is true
\item
{\em and}
{\bf P}$(m)$ is true for all $m < n$.
\end{itemize}
By the Principle of (Finite) Induction, then, we conclude that {\bf
  P}$(n)$ is true for all $n \in \N^+$.
\qed
\end{proof}

\bigskip

We turn now to our second summation, which asserts that each perfect
square of a positive integer, say, $n^2$, is the sum of the first $n$
odd integers, $1, 3, 5, \ldots, 2n-1$.  This proof complements the
constructive proofs of the same result in
Proposition~\ref{thm:squares-odd-integers-Gauss}.

\begin{prop}
\label{thm:squares-odd-integers-induction}
For all $n \in \N^+$,
\[
\sum_{k=1}^n \ (2k-1)
 \ = \ 1 + 3 + 5 + \cdots + (2n-1) \ = \ n^2.
\]
That, is, the $n$th perfect square is the sum of the first $n$ odd
integers.
\end{prop}

\noindent {\em Verification.}
%
For every positive integer $m$, let {\bf P}$(m)$ be the proposition
\[ m^2 \ = \ 1 + 3 + 5 + \cdots + 2m-1. \]
Let us proceed according to the standard format of an inductive
argument.

{\bf 1.} Because $1 \cdot 1 = 1$, proposition {\bf P}$(1)$ is true.

{\bf 2.} Let us assume, for the sake of induction, that proposition
{\bf P}$(m)$ is true for all positive integers strictly smaller than
$n$.  In particular, then, {\bf P}$(n-1)$ is true.

{\bf 3.} Consider now the summation
\[ 1 + 3 + 5 + \cdots + 2n-3 + 2n-1 \]
Because {\bf P}$(n-1)$ is true, we know that
\[ 1 + 3 + 5 + \cdots + 2n-3 + 2n-1 \ = \ (n-1)^2 + 2n-1.  \]
By direct calculation, we see that
\[ (n-1)^2 + 2n-1 \ = \ (n^2 -2n +1) + (2n-1) \ = \ n^2. \]
Because $n$ is an arbitrary positive integer, we conclude that
{\bf P}$(n)$ is true whenever
\begin{itemize}
\item
{\bf P}$(1)$ is true
\item
{\em and}
{\bf P}$(m)$ is true for all $m < n$.
\end{itemize}
By the Principle of (Finite) Induction, then, we conclude that {\bf
  P}$(n)$ is true for all $n \in \N^+$.
\qed



\subsubsection{Making guesses: the method of undetermined coefficients}
\label{sec:undetermined-coefficients}
\index{method of undetermined coefficients}

Proofs by induction, as encountered in
Propositions~\ref{thm:sum-1-to-n-induction}
and~\ref{thm:squares-odd-integers-induction}, are important tools for
{\em verifying} the correctness of alleged results, but induction by
itself is not a tool for {\em discovering} new results.  This section
is devoted to the {\em Method of Undetermined Coefficients}, a tool
that sometimes yields the ``guesses'' that can then be verified via
induction.  We illustrate the method by deriving a formula for the sum
of the first $n$ perfect squares.  Our derivation builds on prior
knowledge of two facts:
\begin{enumerate}
\item
A trivial proof by counting verifies that
\[ \sum_{k=0}^n \ k^0 \ = \ \sum_{k=1}^n \ 1 \ = \ n.  \]
\item
We know from Proposition~\ref{thm:sum-1-to-n-induction} that
\[
\sum_{k=0}^n \ k^1 \ = \ \sum_{k=1}^n \ k \ = \ {1 \over 2}(n^2 + n)
\]
\end{enumerate}
\begin{quote}
It seems to be silly to include the case $n=0$ in our summations,
since that term does not affect the sum, but that case tells us that
the ``constant term'' in all of these polynomials---i.e., the
coefficient of $k^0$---is always $c_0 =0$.
\end{quote}

\begin{prop}
\label{thm:sum-of-squares}
For all $n \in \N$,
\begin{eqnarray}
\nonumber
S^{(2)}_n \ \eqdef \ \sum_{i=1}^n \ i^2 
 & \eqdef &
 1 + 4 + \cdots + (n-1)^2 + n^2 \\
\label{eq:sum-1-to-nsq}
 & = & {1 \over 3} n^3 \ + \ {1 \over 2} n^2 \ + \ {1 \over 6} n
\end{eqnarray}
\end{prop}
\index{formula for the sum of the first $n$ squares}

The target quantity $S^{(2)}_n$ in the proposition is often expressed
in a more aesthetic form:
\[ S^{(2)}_n \ = \
{1 \over 6} n (n+1)(2n+1) \ = \
\frac{2n+1}{3} \cdot {n \choose 2}.
\]

\begin{proof}
Since summing $0$th powers thus gives us a degree-$1$ polynomial, and
summing $1$st powers gives us a degree-$2$ polynomial, it is not
unreasonable to guess that summing $2$nd powers would give us a
degree-$3$ polynomial.  This turns out to be a good guess!  To prove
this assertion, we must determine values for constants $c_1, c_2, c_3$
such that
\begin{equation}
\label{eq:symbolic-cubic}
\sum_{i=0}^n \ k^2 \ = \ c_3 n^3 + c_2 n^2 + c_1 n.
\end{equation}
(Including the case $n=0$ leaves us with {\em three} constants to
determine rather than four: we know that the coefficient of $k^0$ is
$c_0 =0$.)

We begin our determination of the constants $c_1, c_2, c_3$ by
instantiating the symbolic polynomial in (\ref{eq:symbolic-cubic}) at
the smallest three values of $n$.  Any three values will work; using
the {\em smallest} ones simplifies our calculations.  These
instantiations leaves us with the following system of linear
equations.  (The summations in (\ref{eq:undetermined}) indicate where
each linear equation in the system comes from.)
\begin{equation}
\label{eq:undetermined}
\begin{array}{lccccccccc}
1. &
{\displaystyle \sum_{i=0}^1 \ k^2}
   & = & c_3    & + & c_2   & + & c_1   & = & 1 \\
2. &
{\displaystyle \sum_{i=0}^2 \ k^2}
   & = & 8 c_3  & + & 4 c_2 & + & 2 c_1 & = & 5 \\
3. &
{\displaystyle \sum_{i=0}^3 \ k^2}
   & = & 27 c_3 & + & 9 c_2 & + & 3 c_1 & = & 14
\end{array}
\end{equation}

We use a form of the {\it Gaussian elimination}
algorithm\footnote{This algorithm is defined and validated in sources
  such as \cite{CLRS}.}~to solve the system by ``eliminating
variables.''  First, we rewrite equation 1 as
\[ 8 c_3 + 8 c_2 + 8 c_1 \ = \ 8 \]
and substract equation 2 from it, thereby obtaining the $2$-variable
equation
\begin{equation}
\label{eq:step1}
4c_2 + 6 c_1 \ = \ 3.
\end{equation}
We perform a similar calculation based on equations 1 and 3 in
system (\ref{eq:undetermined}):  We rewrite equation 1 as
\[ 27 c_3 + 27 c_2 + 27 c_1 \ = \ 27 \]
and substract equation 3 from it, thereby obtaining the $2$-variable
equation
\begin{equation}
\label{eq:step2}
18 c_2 + 24 c_1 \ = \ 13.
\end{equation}
We now rewrite equations (\ref{eq:step1}) and (\ref{eq:step2}) to
obtain the simplified system
\[
\begin{array}{ccccc}
72 c_2 & + & 108 c_1 & = & 54 \\
72 c_2 & + &  96 c_1 & = & 52
\end{array}
\]
We now see, via subtraction, that
\[ 12 c_1 \ = \ 2 \]
or, equivalently,
\begin{equation}
\label{eq:valueof-c1}
c_1 \ = \ 1/6.
\end{equation}

Now that we know the value of $c_1$, we can make the indicated
substitution and further simplify system (\ref{eq:undetermined}).
Since we now have only two variables, we can also eliminate any one
of the three equations in (\ref{eq:undetermined}).  We eliminate
equation 3 in the system in order to simplify our calculations from
this point on.  We now have the system
\begin{equation}
\label{eq:undetermined-step2}
\begin{array}{lccccc}
1. &
c_3  & + & c_2   & = & 5/6 \\
2. &
8c_3 & + & 4 c_2 & = & 14/3 
\end{array}
\end{equation}
We rewrite equation 1 in (\ref{eq:undetermined-step2}) as
\[ 8 c_3 + 8 c_2 \ = \ 20/3 \]
and subtract equation 2 from this version of equation 1.  We thereby
discover that
\[ 4 c_2 \ = \ 2, \]
or, equivalently,
\begin{equation}
\label{eq:valueof-c2}
c_2 \ = \ 1/2.
\end{equation}

Using the values we have discovered for $c_1$ and $c_2$, in equations
(\ref{eq:valueof-c1}) and (\ref{eq:valueof-c2}), respectively, we
finally use equation 1 of system (\ref{eq:undetermined}) to determine
the value of $c_3$:
\begin{equation}
\label{eq:valueof-c3}
c_3 \ = \ 1 - \ 1/2 \ - \ 1/6 \ = \ 1/3.
\end{equation}
We have, thus, derived equation~(\ref{eq:sum-1-to-nsq}).

\medskip

The careful reader will note that we are not really done yet.  We
have derived our expression for $S^{(2)}_n$ under the as-yet
unjustified assumption that $S^{(2)}_n$ really is a cubic (i.e.,
degree-$3$) polynomial.  What we need now is an induction that will
verify our result.  With our previous illustrations as models, we
shall leave this final task to the reader.

\noindent
***************** \\
IS THIS OK? \\
***************** 
\qed
\end{proof}

With more (calculational) work, but no new (mathematical) ideas, one
can derive explicit expressions for the sum of the first $n$ $k$th
powers, i.e., the quantity $S^{(k)}_n$, for any positive integer $k$.


\subsection{Proof by Contradiction}
\label{sec:Contradiction}
\index{proof by contradiction}

\subsubsection{The Proof Technique}
\label{sec:contradiction-technique}
\index{proof by contradiction!technique}

The basic principle that underlies proof by contradiction is that the
following {\em metamathematical} assertions are logically equivalent.
\begin{itemize}
\item
Proposition $P$ {\sc implies} Proposition $Q$
\item
Proposition $\sim Q$ {\sc implies} Proposition $\sim P$
\end{itemize}

As with many of these {\em metamathematical} principles, there is a
corresponding {\em mathematical} equivalence, in this case,
Proposition~\ref{thm:contraposition}.
\[ [P \Rightarrow Q] \ \equiv \ [(\sim Q) \Rightarrow (\sim P)] \]


\subsubsection{Sample Proofs}
\label{sec:sample-contradictions}
\index{proof by contradiction!sample proofs}

\addcontentsline{toc}{paragraph}{A. There are infinitely many primes}
\noindent{\small\sf A. There are infinitely many primes.}
%
The following result is traditionally attributed to the Greek
mathematician Euclid,
\index{Euclid}
one of the patriarchs of mathematics.

\begin{prop}
\label{thm:Primes-infinite}
There are infinitely many prime numbers.
\end{prop}

\begin{proof}
Let us assume, contrarily, that there are only finitely many primes.
Say, in particular, that the following $r$-element sequence enumerates
all (and only) primes, in order of magnitude:

$\mbox{\bf Prime-Numbers} \ = \ 
\langle P_1, \ P_2, \ \ldots, \ P_r \rangle$

\noindent where
\begin{itemize}
\item
$P_1 = 2$
\item
$P_2 = 3$
\item
$P_i < P_{i+1}$ for all $i \in \{1, 2, \ldots, r-1\}$.
\end{itemize}

We verify the {\em falseness} of the alleged completeness of the sequence
{\bf Prime-Numbers} by analyzing the positive integer
\[ n \ = \ 1 + \prod_{i=1}^r \ P_i \ = \ 1 \ + \ 
\left(P_1 \cdot P_2 \cdot \cdots \cdot P_r \right).
\]

We make three crucial observations.

\begin{enumerate}
\item
We note first that {\em the number $n$ is not divisible by any prime
number  in the sequence {\bf Prime-Numbers}.}

To see this, note that for each $P_k$ in the sequence,
\[
n / P_k \ = \ \frac{1}{P_k} \ + \ \prod_{i \neq k} \ P_i .
\]
Because $P_k \geq 2$, we see that $n / P_k$ obeys the inequalities
\[
\prod_{i \neq k} \ P_i \ < \ n/P_k \ < \ 1 + \prod_{i \neq k} \ P_i.
\] 
The discreteness of the set $\Z$---see
Section~\ref{sec:integers}.A---implies that $n / P_k$ is not an
integer, because it lies strictly between two adjacent integers.

\item
We note next that, because of assertion 1, if the sequence {\bf
  Prime-Numbers} actually contained {\em all} of the prime numbers,
then we would have to conclude that {\em the number $n$ is not
  divisible by any prime number.}

\item
Finally, we remark that the Fundamental Theorem of Arithmetic
(Theorem~\ref{thm:Fund-Thm-Arith}) implies that {\em every integer is
  divisible by (at least one) prime number}.
\end{enumerate}

We have a chain of assertions that lead to a mutual inconsistency: on
the one hand, the integer $n$ has no prime-integer divisor; on the
other hand, no integer can fail to have a prime-integer divisor!  Let
us analyze how we arrived at this uncomfortable place.
\begin{itemize}
\item
At the front end of this uncomfortable string of assertions we have
the assumption that there are only finitely many prime numbers.  We
have (as yet) no substantiation for this assertion.
\item
At the back end of this uncomfortable string of assertions we have
the ({\em rock solid}) Fundamental Theorem of Arithmetic
(Theorem~\ref{thm:Fund-Thm-Arith}).
\item
In between these two assertions we have a sequence of assertions, each
of which follows from its predecessors via irrefutable rules of
inference.
\end{itemize}
It follows that the {\em only} brick in this edifice that could be
faulty---i.e., the only assertion that could be false---is the
assumption that there are only finitely many prime numbers.  Since
this assumption leads to an inconsistent set of assertions, we must
conclude that the assumption is false!  We conclude from this
classical proof by contradiction that there are infinitely many prime
numbers.  \qed
\end{proof}


\subsection{Proofs via the Pigeonhole Principle}
\label{sec:pigeonhole}

The proof technique we discuss now builds on an observation that is
almost embarrassingly obvious---yet its simplicity is exceeded by its
importance as a source of strikingly surprising results.

\subsubsection{The Proof Technique}

The technique, known variously as {\it the pigeonhole principle}
\index{pigeonhole principle}
or {\it Dirichlet's Box Principle}
\index{Dirichlet's Box Principle}
(after the French mathematician Peter Gustav Lejeune Dirichlet),
\index{Dirichlet, Peter Gustav Lejeune}
exploits the fact that if one has $n$ objects (say, pigeons) and $m <
n$ boxes (they're the pigeonholes), then any way of putting pigeons
into boxes must place at least two pigeons into the same box.


\subsubsection{Sample (Fun) Applications/Proofs}
\label{sec:pigeon-apps}

\addcontentsline{toc}{paragraph}{A. Choosing a pair of matching socks}
\noindent{\small\sf A. Choosing a pair of matching socks.} 
%
You have $n$ pairs of socks, the socks in each pair having a distinct
color (one pair of red socks, one pair of blue socks, \ldots).  Since
you wake up ``very slowly'', you want to grab some number of unpaired
socks that is cerain to yield at least one pair of smame-color socks.
Clearly, if you grab any $n+1$ socks (the pigeons), the pigeonhole
principle guarantees that you have at least one monochromatic pair,
because there are only $n$ distinct sock-colors (the boxes).

\medskip

\addcontentsline{toc}{paragraph}{B. Finding birthday-mates}
\noindent{\small\sf B. Finding birthday-mates.}
%
You are attending a conference and wander into a lecture that has 367
attendees (including you).  It is certain that at least two attendees
share the same birthday: there are 366 possible birthdays (the boxes
for a leap year) and 367 birthday-possessors (the pigeons).

\medskip

\addcontentsline{toc}{paragraph}{C. On meeting new people}
\noindent {\small\sf C. On meeting new people.}
%
You are attending a cocktail party that is populated by $n$ couples.
In order to create a warm atmosphere, the host requests that each
attendee shake the hand of every attendee that he or she does not
know.  The pigeonhole principle guarantees that some two attendees
shake the same number of hands.  To wit, the number of people that
each attendee {\em does not know} belongs to the set $\{ 0, 1, \ldots,
2n-2 \}$, because each person knows him/herself and his/her partner.
Because there are $2n$ handshakers (the pigeons) and $2n-1$ numbers of
hands to shake (the boxes), some two shakers must shake the same
numbers of hands.

\medskip

\addcontentsline{toc}{paragraph}{D. Friends and strangers at a party}
\noindent{\small\sf D. Friends and strangers at a party.}
%
We turn now to a somewhat more surprising result that can be proved
using the pigeonhole principle.  While we phrase the result in
anthropomorphic, ``homely'', terms, its formal statement identifies it
as a genre of ``unavoidable subgraph''
\index{unavoidable subgraph phenomena}
%
phenomenon within the theory of {\it graphs}.
\begin{quote}
Graphs are an immensely important mathematical construct that models
myriad situations that involve objects (possibly people) and
interrelationships between pairs of objects.
Chapter~\ref{Ch:Graphs-Trees} is devoted to studying graphs and the
situations they can be used to model---including the problem discussed
here.
\end{quote}
Here is the ``homely'' version of the {\it Friends and Strangers} problem.
\index{friends and strangers problem}

\begin{prop}
\label{thm:triangle-cotriangles}
In any gathering of six people, at least one of the following
assertions is true.

\noindent {\rm A.}
There is a group of three people who know each other.

\noindent {\rm B.}
There is a group of three people none of whom knows either of the
others.
\end{prop}

\begin{proof}
Let the gathering consist of six indistinguishable people, named
$P_1$, $P_2$, $P_3$, $P_4$, $P_5$, $P_6$.  Focus on an arbitrary
person, say $P_5$.  (This choice ``sounds'' more arbitrary than
$P_1$---but, of course, is not.)  Now, there are $5$ people, namely,
$P_1$, $P_2$, $P_3$, $P_4$, $P_6$, each of whom $P_5$ either {\em
  knows} or {\em doesn't know}.

Clearly, some $3$ of these $5$ people ``lie on the same side of the
{\em know/don't-know} fence.''  This follows from the pigeonhole
principle: we have {\em two} boxes ({\em know} and {\em doesn't know})
and {\em five} pigeons (the people $P_1$, $P_2$, $P_3$, $P_4$, $P_6$).
Any way of putting the pigeons into the boxes will place three people
into one of the boxes.

Say, with no loss of generality, that $P_5$ {\em knows} $P_1$, $P_2$,
$P_3$.
\begin{quote}\index{``with no loss of generality'': meaning}
Why can we claim that the selected situation--- ``$P_5$ {\em knows}
$P_1$, $P_2$, $P_3$''---can be assumed ``with no loss of generality''?
One should {\em always} ask this question about such a claim!  In the
current case, the claim follows from the following facts.

(a) The names that we use to refer to the six assembled people are
just for our expository benefit.  The names carry no inherent meaning
related to the {\it Friends and Strangers} problem.  You can repeat
our argument while choosing arbitrary replacements for $P_1$, $P_2$,
$P_3$, $P_5$, with no change to the logical outcome.

You can also interchange the {\em know} and {\em don't-know} labels.
The underlying logic will not change, although the conclusions
regarding options A and B in the statement of the proposition will
clearly ``flip''.
\end{quote}

Having decided that $P_5$ {\em knows} $P_1$, $P_2$, and $P_3$, we now
consider the implications of the possible relations between each of
the three pairs of people chosen from $\{P_1, P_2, P_3\}$.  There
are two logical possibilities.
\begin{itemize}
\item
Some two of $P_1$, $P_2$, $P_3$ know each other---say, with no loss of
generality, $P_1$ and $P_2$.  In this case, $P_1$, $P_2$, and $P_5$
form a trio of people who know one another (option A in the statement
of the proposition).
\item
No two of $P_1$, $P_2$, $P_3$ know each other.  In this case, $P_1$,
$P_2$, and $P_3$ form a trio of people none of whom knows either of
the others (option B in the statement of the proposition).
\end{itemize}
This disjunction completes the proof.

We close the proof by noting that nothing we have stated precludes the
possibility that {\em both} option A {\em and} option B are true!  \qed
\end{proof}


\section{Reasoning via Mathematical Analysis}
\label{sec:analysis}

\subsection{Analyses via Linear Recurrences}
\label{sec:linear-recurrences-1}

\begin{theorem}[The Master Theorem for Linear Recurrences]
\label{thm:master-thm-1}
\index{The Master Theorem for Linear Recurrences}
Let the function $F$ be specified by the following linear recurrence.
\begin{equation}
\label{eq:Lin-Recur:start-1}
F(n) \ = \ \left\{
\begin{array}{cl}
a F(n/b) + c & \mbox{for } n \geq b \\
c & \mbox{for } n < b
\end{array}
\right.
\end{equation}
Then the value of $F$ on any argument $n$ is given by
\begin{equation}
\label{eq:Lin-Recur:solve-1}
\begin{array}{lclll}
F(n) & = & (1 + \log_b n)c &  & \mbox{if } a=1 \\
     &   &                 &  & \\
     & = &
  {\displaystyle
  \frac{1-a^{\log_b n}}{1-a} \ \approx \ \frac{1}{1-a}
  }
 &  & \mbox{if } a<1 \\
    &   &                  & & \\
    & = &
  {\displaystyle
\frac{a^{\log_b n} -1}{a-1}
  }
 & & \mbox{if } a>1
\end{array}
\end{equation}
\end{theorem}




\subsection{Asymptotics}
\label{sec:asymptotics}

{\em Asymptotics} can be viewed as a language and a system of
reasoning that allow one to talk in a {\em qualitative} voice about
{\em quantitative} topics.  We thereby generalize to arbitrary growth
functions terms such as ``linear'', ``quadratic'', ``exponential'',
and ``logarithmic''.

Such a language and system are indispensable if one needs to reason
about computational topics over a range of situations, such as a range
(``all existing''?)  computer architectures and software systems.  As
two simple examples: (1) Carry-ripple adders perform additions in time
linear in the lengths $n$ of the summands (measured in number of bits)
no matter what these lengths are. (2) Comparison-based sorting
algorithms can sort lists of $n$ keys in time proportional to $n \log
n$, but no faster---where the base of the logarithm depends on the
characteristics of the computing platform.  More precise versions of
the preceding statements require specication of the number $n$ and
other details, possibly down to the clock speeds of the host
computer's circuitry.

\subsubsection{The language of asymptotics}

The language of asymptotics, which has its origins in the field of
Number Theory in the late 19th century, builds on the following
terminology, which is likely what one would cover in an early
undergraduate course.  More advanced aspects of the language would
likely by beyond the needs of most students of computing, aside from
specialists in advanced courses.  The basics of the language build on
three primitive notations and notions.  Standard sources, such as any
text on algorithm design and analysis, flesh out the following ideas.
\begin{itemize}
\item
{\em The big-O notation}.
%
The assertion $f(x) = O(g(x))$ says, intuitively, that the function
$f$ grows no faster than function $g$.  It is, thus, the asymptotic
analogue of ``less than''.

Formally:
$f(x) = O(g(x))$

means

$(\exists c >0)(\exists x^{\#})(\forall x > x^{\#})
[f(x) \leq c \cdot g(x)]$

\item
{\em The big-$\Omega$ notation}.
%
The assertion $f(x) = \Omega(g(x))$ says, intuitively, that the
function $f$ grows at least as fast as function $g$.  It is, thus, the
asymptotic analogue of ``greater than''.

Formally:
$f(x) = \Omega(g(x))$

means

$(\exists c >0)(\exists x^{\#})(\forall x > x^{\#})
[f(x) \geq c \cdot  g(x)]$ \\

\item
{\em The big-$\Theta$ notation}.
%
The assertion $f(n) = \Theta(g(n))$ says, intuitively, that the
function $f$ grows at the same rate as does function $g$.  It is,
thus, the asymptotic analogue of ``equal to''.

Formally:
$f(x) = \Theta(g(x))$

means

$(\exists c_1 >0)(\exists c_2 >0)(\exists x^{\#})(\forall x > x^{\#})
[c_1 \cdot g(x) \leq f(x) \leq c_2 \cdot  g(x)]$
\end{itemize}
One renders the preceding intuitive explanations precise by pointing
out that the three specifies relations ($a$) take hold {\em
  eventually}, i.e., only for large arguments to the functions $f$ and
$g$, and ($b$) hold up to an unspecified constant of proportionality.

\ignore{*********
\subsubsection{Getting formal}

{\small\sf Big-$O$, Big-$\Omega$, and Big-$\Theta$ notation}.%
It is convenient to have terminology and a notation that allows us to
talk about the rate of growth of one function as measured by the rate
of growth of another.  We are interested in the exact growth rate, as
well as upper and lower bounds on the growth rate.  We do have
appropriate such language for certain rates of growth.  We can talk,
for instance, about a linear growth rate or a quadratic rate or an
exponential rate, to name just a few---and we get the desired bounds
using the prefixes ``sub'' or ``super,'' as in ``subexponential'' and
``superlinear''---but our repertoire of such terms is quite limited.
Mathematicians working in the theory of numbers in the late nineteenth
century established a notation that gives us an unlimited repertoire
of descriptors for growth rates, via what has come to be called the
big-$O$, big-$\Omega$, and big-$\Theta$ notations, which are
collectively sometimes called {\em asymptotic notation}.\index{asymptotic notation}
*******}

\subsubsection{The ``uncertainties'' in asymptotic relationships}

The formal definitions of all three of our asymptotic relationships
are bracketed by two important quantifiers:
\[ ``(\exists c >0)'' \ \ \ \mbox{ and } 
 ``(\forall x > x^{\#})''.
\]
The former, {\em uncertain-size} quantifier, asserts that asymptotic
notions describe functional behavior ``in the large''.  Thus, in
common with more common qualitative descriptors of quantitative growth
such as linear, quadratic, cubic, quartic, exponential, logarithmic,
etc., asymptotic relationships give no infomation about constants of
proportionality.  {\em We are not saying that constant factors do not
  matter!  We are, rather, saying that we want to discuss growth
  patterns \underline{in the large}.}

The latter, {\em uncertain-time} quantifier asserts that asymptotic
relationships between functions are promised to hold only
``eventually'', i.e., ``for sufficiently large values of the argument
$x$''.  Therefore, in particular, asymptotic notions cannot be
employed to discuss or analyze quantities that can never grow beyond a
fixed finite value.  The fact that all instances of a quantity
throughout history have been below $N$ is immaterial, as long as it is
conceivable that an instance larger than $N$ could appear at some time
in the future.

These quantifiers in particular distinguishes claims of asymptotic
relationship from the more familiar definite inequalities such as
``$f(x) \leq g(x)$'' or $f(x) \geq 7 \cdot g(x)$.  In fact, it is
often easier to think about our three asymptotic bounding assertions
as establishing {\em envelopes} for $f(x)$:
\begin{itemize}
\item
Say that $f(x) = O(g(x))$.  If one draws the graphs of the functions
$f(x)$ and $c \cdot g(x)$, then as one traces the graphs with
increasing values of $x$, one eventually reaches a point $x^{\#}$
beyond which the graph of $f(x)$ never enters the territory {\em
  above} the graph of $c \cdot g(x)$.
\item
Say that $f(x) = \Omega(g(x))$.  This situation is the up-down mirror
image of the preceding one: just replace the highlighted ``{\em
above}'' with ``{\em below}.''
\item
Say that $f(x) = \Theta(g(x))$.  We now have a two-sided envelope:
beyond $x^{\#}$, the graph of $f(x)$ never enters the territory {\em
  above} the graph of $c_1 \cdot g(x)$ and never enters the territory
{\em below} the graph of $c_2 \cdot g(x)$.
\end{itemize}
In addition to allowing one to make familiar growth-rate comparisons
such as ``$n^{14} = O(n^{15})$'' and ``$1.001^n = \Omega(n^{1000})$,''
we can now also make assertions such as ``$\sin x = \Theta(1)$,''
which are much clumsier to explain in words.

\medskip

\noindent {\bf Beyond the big letters.}
%
There are ``small''-letter analogues of the preceding ``big''-letter
asymptotic notations, but they are only rarely encountered in
discourse about real computations (although they do arise in the
analysis of algorithms).

\subsubsection{Inescapable complications}

The story we have told thus far is covered in many sources and
courses.  Two complications to the story are covered less faithfully,
although lacking them, one cannot perform cogent asymptotic
reasoning.  Both complications involve the notion of {\em uniformity}.

\noindent
{\bf 1.}
{\em Multiple functions}.
%
Say that we have four functions, $f, g, h, k$, and we know that both
\[ f(n) = O(g(n)) \ \ \mbox{ and } \ \ h(n) = O(k(n)) \]
It is intuitive that
\[ f(n) + h(n) = O(g(n) + k(n)) \]
--- but is it true?

In short, the answer is YES, but verifying that requires a bit of
subtlety, because, absent hitherto undisclosed information, the
proportionality constants $c_{f,g}$ and $N_{f,g}$ that witness the
big-$O$ relationship between functions $f$ and $g$ have no connection
with the constants, $c_{h,k}, N_{h,k}$ that witness the analogous
relationship between functions $h$ and $k$.  Therefore, in order to
verify the posited relationship between functions $f + h$ and $g + k$,
one much find witnessing constants $c_{f+h, g+k}$ and $N_{f+h,g+k}$.
Of course, this task requires only elementary reasoning and
manipulation --- but it must be done!

{\bf 2.}
{\em Multivariate functions}.
%
Finally, we discuss the scenario that almost automatically accompanies
the transition from a focus on sequential, single-agent computing to a
focus on PDC.  Within this broadened context, most functions that
describe a system have one or more variables that describe the
computing system --- its number of processors or of agents or the
sizes of its memory modules or the communication-radii of its
transponders or \ldots, in addition to the one or more variables that
describe the data that the system is processing.  Within such
scenarios, every assertion of an asymptotic relationship, of the form
\[ f(\vec{m}; \vec{n}) = O(g(\vec{m}; \vec{n})) \]
must explicitly specify the following information:
\begin{itemize}
\item
which variables can grow without bound;
\item
among such unbounded variables, which participate in the posited
asymptotic relation;
\item
for each participating unbounded variable $x$, what are the constants
$c_x$ and $N_x$ that witness the posited asymptotic relationship(s).
\end{itemize}

Clearly the complexity of cogent asymptotic reasoning --- hence also
the complexity of teaching about such reasoning --- gets much more
complicated in the multivariate settings engendered by PDC.  But, the
benefits of being able to reason qualitatively about the quantitative
aspects of computing increase at least commensurately!



